---
title: 'SYST 664: Assignment 6'
author: "Jake Whalen"
date: "March 20, 2018"
output: 
  html_document: 
    toc: yes
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F)
```

```{r packages}
source("src/_setup.R")
```

# Problem 1
Concentrations of the pollutants aldrin and hexachlorobenzene (HCB) in nanograms per liter were measured in ten surface water samples, ten mid-depth water samples, and ten bottom samples from the Wolf River in Tennessee.
The samples were taken downstream from an abandoned dump site previously used by the pesticide industry.
The full data set can be found at http://www.biostat.umn.edu/~lynn/iid/wolf.river.dat.
The code below reads this data set into R.

```{r, echo=T}

```

For this problem, we consider only HCB measurements taken at the bottom and the surface.
The question of interest is whether the distribution of HCB concentration depends on the depth at which the measurement was taken.
The data for this problem are given below.