---
title: "SYST 664: HW Assignment 3"
author: "Jake Whalen"
date: "February 12, 2018"
output: word_document
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(echo = F, 
                      warning = F, 
                      fig.align = 'center', 
                      fig.width=6, 
                      fig.height=3)
```

```{r}
source("03_setup.R")
source("03_problem_01.R")
source("03_problem_02.R")
```


# Problem 1

### Part A

```{r}
p1.a.plot
```

### Part B

```{r}
p1.b.plot
```

### Part C

```{r}
p1.c.plot
```

### Part D

```{r}
p1.d.plot
```

### Part E
##### What do the results in parts a-d say about changes in farming in the population from which these data are drawn?

Marginal probability of farming being an individuals occupation drops from fathers (`r filter(p1.a, father == 'farm')$Marginal`) to sons (`r filter(p1.b, son == 'farm')$Marginal`).
If the father was a farmer then the son has a `r percent(filter(p1.c, son == 'farm')$Conditional)` probability of also taking up the farming occupation.
If the son was a farmer then the father was a farmer with a `r percent(filter(p1.d, father == 'farm')$Conditional)` probability.
Thus while farming may not be passed down at a high rate between these generations it also is not an occupation that the younger generation is flocking to.
If the son is a farmer it is unlikely their father had a non-farming job.

# Problem 2

### Part A

```{r}
qqplot(exp.quant, interArrival, 
       main='Exponential Q-Q Plot of Car Arrival Times', 
       xlab='Theoretical Exponential Quantiles',ylab='Empirical Quantiles')
lines(exp.quant, exp.quant *mean(interArrival))
```

Do you think an exponential distribution provides a good model for the interarrival times? 

Justify your answer.

### Part B

```{r}
p2.b.plot
```

Do you think a Poisson distribution provides a good model for the count data? 

Justify your answer.

### Part C

```{r, fig.height=6}
grid.arrange(p2.c.plot.1, 
             p2.c.plot.2, 
             nrow=2,
             left='Posterior Probability',
             bottom='# of Cars Passing Point (in 15 Second Blocks)')
```

Find the posterior mean, standard deviation, median and 95  percentile of Lambda given the observations. 

Describe what your results mean in terms of traffic on this motorway.